Three Problems for Sophistication

### 2015-01-18 Facticity/Facticity/paper.tex
...

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{\tn{soph}}
\newcommand{\num}[1]{#1_1, #1_2, \ldots}

@@

\begin{abstract}
Kolmogorov complexity is a sound measure of the amount of information in a binary string, but it does not distinguish between the information that expresses the structural properties of the data, and noise. Kolmogorov's definition of the \emph{structure function} in his lecture Tallinn in 1973 was the first attempt to separate meaningful information from noise, but many alternative definitions of the \emph{sophistication}, or the amount of meaningful information in the data, have subsequently been proposed. This indicates that the idea is considered worth pursuing by multiple distinguished scholars, but at the same time it shows that the issue has never been completely satisfactorily resolved.

In this paper we describe three fundamental problems that arise with all existing proposals, showing each of them to be \emph{unsound}. Based on these results we we put forward the view that the problem is fundamental: while many nontrivial and useful structural properties of binary strings can be detected without a priori assumptions, it may be impossible to objectively quantify its sophistication.
\end{abstract}

@@

Kolmogorov complexity breaks away from the tradition in statistics and information theory by allowing the analysis of \emph{individual} binary sequences rather than studying probability distributions. This is important because for many binary strings, such as a binary encoding of a novel, or a holiday snapshot, or the fossil record, it is impossible to determine what distribution generates such objects. There is, however, a nagging discrepancy between Kolmogorov complexity and what most people whould intuitively call ``complex''. For example, a sequence of a million coin flips is exceedingly unlikely to have a Kolmogorov complexity of much less than one million bits, even though there is nothing complex about flipping a lot of coins---it's just hard work. For this reason, many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data. This tradition started with a lecture by Kolmogorov in Talinn, Estonia in 1973, but many alternative definitions of meaningful information, or \emph{sophistication}, were proposed, see the related work section below. In this paper, we investigate three problems with these definitions, and show that they are unsatisfactory. We believe the problem to be fundamental, and provide several intuitive arguments why these attempts to define sophistication must fail.

The Kolmogorov complexity of a binary string $x$ is, roughly, the length of the shortest computer program that will output $x$. This length depends on the used programming language. However, the Invariance Theorem states that the choice between any two Turing complete languages affects the Kolmogorov complexity by at most a constant term: if $K^L(x)$ denotes Kolmogorov complexity of $x$ with respect to language $L$, then $|K^L(x)-K^M(x)|<c$ for some constant $c$ that depends on $L$ and $M$, but \emph{not on $x$}. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings. The invariance theorem allows us to fix a single reference programming language and omit it from the superscript.

@@

\begin{enumerate}
\item Being an information measure, the sophistication should count the number of bits required for an effective description of the structural properties of a binary string.
\item An analogue of the Invariance Theorem should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the used programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input sequence $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language, even for inputs of very high Kolmogorov complexity. Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that the $K(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity, and the sophistication would incorporate the amount of noise in the data just like Kolmogorov complexity does.
\end{enumerate}

For most proposed definitions of sophistication, we can prove that they fail one or more of the attributes above. Vit\'anyi's definition \cite{TODO} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only extremely deep strings (i.e., strings that require an enormous amount of processing to construct) yield a large sophistication, while intuitively it seems there must be objects that contain a lot of structure all of which can be described without much calculation. Moreover in our discussion section we provide arguments that suggest that it should be possible to generalise our negative results to show that Vit\'anyi's sophistication is not invariant.

A valid definition of sophistication must contend with the following issues. First, while all definitions are based on a two-part encoding of the data sequence, some researchers use Kolmogorov complexity as a \emph{building block} of their proposed code, while others look for the two parts of the code \emph{inside} the Kolmgorov complexity: when the definition of Kolmogorov complexity is opened up, it turns out to consist of two parts itself. In Section~\ref{sec:nickname}, we show how this second approach sets us up to suffer from to the so-called \emph{nickname problem}: the sophistication becomes highly dependent on the chosen programming language. This immediately breaks invariance, unless additional requirements are made on the language. We also show that there do exist languages (efficient enumerations of the partial recursive functions) that address this issue.

The second issue is that of \emph{overfitting}, which we consider in Section~\ref{sec:overfitting}. Overfitting is a common problem in the statistical literature, that refers to the tendency of some model selection procedures to select a very complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. For example, in a polynomial regression setting, a higher order polynomial can always provide better fit to a data set. Therefore, even if the data are in fact noisy samples from a parabola, naive model selection methods may select a much higher order polynomial that provides a worse description of the trend in the data. The same can occur for the sophistication. We somehow need to make sure that the two-part representation that determines the sophistication does not store noise in the ``structure'' part of the code. In statistics, this is typically addressed by penalising complex models, e.g. using the BIC criterion \cite{TODO}. However, such penalisations tend to break the delicate balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting (Section~\ref{sec:underfitting}) is the opposite problem, where the selected model is simple, but fails to represent part of the structure in the data. It is an especially serious problem for sophistication because the considered models for sophistication are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such interpreters can generate \emph{any} data sequence with only a short input, and if complex models are penalized to avoid overfitting, then the problem becomes to make sure that interpreters are not \emph{always} preferred for complex data. This problem is widely recognised, and is sometimes referred to as the ``collapse'' of the structure function \cite{}. The usual workaround is to restrict the set of considered models such that interpreters are no longer allowed. For example, the model class may be restricted to the set of all total recursive functions (only programs that never hang) rather than all partial recursive functions (all programs). While this patches the ``collapse'' problem to an extent, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, in the discussion in Section~\ref{sec:discussion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as unlikely to capture all structure in the data, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable models of the data generating mechanism. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that sometimes the data allows for multiple very different interpretations, and that there is no such thing as sophistication.

\subsection{Related work}
The idea of separating structure from noise originated with Kolmogorov's lecture in Tallinn, Estonia in 1973 about the \emph{structure function},. This work was never published, but summarized by Cover \cite{cover1985kolmogorov}.  In this theory, the two-part description of the data consists of a program that generates a finite set, and an index of the original object in the set. The length of the program then measures the structure in the data. This approach was discussed more extensively, and generalised, by Vereshchagin, G\'acs and Vit\'anyi \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}).

@@

In 1996 Gell-Mann and Lloyd introduced `Effective complexity' \cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that fits the mold of sophistication.

Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel to separate candidate two-part descriptions from those deemed to be too inefficient, is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth.

Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. Adriaans also identifies the nickname problem discussed in Section~\ref{sec:nickname}.

@@

We will generalizes the main proposals for sophistication into a single, consistent notation. The unifying principle is this: we start with a class of models, we consider the space of two-part descriptions using these models, we select the set of representations with the shortest code-lengths and we select, by some method, the best model from this set.

@@

As noted in the introduction, the Kolmogorov complexity is invariant to the choice of description language. Formally, our description language creates a numbering of the partial computable functions.

\begin{definition}[Numberings]
A numbering is an enumeration of $\C$, denoted as $\phi_1, \phi_2, \ldots$ or by a function $g: \N \to \C$. We fix one canonical numbering $g^\circ$, which works as an effective description, that is for any $i \in \N, y\in \B$, we can effectively compute $(g^\circ(i))(y)$. We call a numbering $g$ \emph{acceptable} if there exist total, computable functions $a, b: \N \to \N$ as that for all $i$ $g^\circ(a(i)) = h(i)$ and  $g^\circ(i) = h(b(i))$.
\end{definition}

When we say that Kolmogorov complexity is invariant, we mean that changing the canonical acceptable numbering changes the Kolmogorov complexity by at most a constant.

@@

\begin{definition}[Complexity]
Let $\phi_1,\phi_2,\ldots$ be the canonical acceptable enumeration of $\C$. Let $\Pi$ be a computable subset of the natural numbers such that  $\{\phi_i:i\in\Pi\}$ corresponfs to the set of all computable prefix functions. Now define Kolmogorov complexity by
\begin{align*}
C(x)&=\min\{|\bar\imath y|:\phi_i(y)=x\},\\
K(x)&=\min\{|\bar\imath y|:\phi_i(y)=x, i\in\Pi\}.\\
\end{align*}
\end{definition}
This definition of $K$ is uncommon. It is equivalent to the standard definition as used in \cite{li1993introduction}, up to a constant, but this form is more convenient for defining the $K$-complexity of non-prefix functions (see Definition~\ref{definition:model-complexity} below).

We now have the basic tools with which we set up our generic sophistication function. We start with the class of allowed models:

\begin{definition}[Models and model classes] A \emph{model class} is a set of models. We distinguish four important model classes:
  \begin{itemize}
  \item $\C$ is the set of all partial computable functions,
  \item $\K\subset\C$ is the set of all prefix-free
    functions,
  \item $\T\subset\C$ is the set of all total functions,
  \item $\P\subset\K$ is the set of all complete, prefix-free functions,
  \item $\F\subset\P$ contains, for every finite set $S$, a computable
    surjective function mapping the binary sequences of length
    $\lceil\log S\rceil$ onto $S$. In other words, this is the model class of finite sets, with uniform distributions.
  \end{itemize}
\end{definition}

We note that the model classes $\T$ and $\P$ are unusual in that they are not enumerable.

\begin{definition}[Representations and code-lengths]
A pair $(m,y)$ is a \emph{representation} for $x$ if $m$ is a model and $m(y)=x$.

Define the following code-length functions for the representations:
\begin{itemize}
\item $L(m)$ is the length of a self-delimiting description of
  the model $m$, i.e. it is the length function of a prefix code for models. Each variant of sophistication must supply such a function.
\item $L^m(x)=L(m)+\min\{|y|:m(y)=x\}$ measures the number of bits required to describe both the model $m$ and the data $x$ using the model.
\item $L^\M(x)=\min\{L^m(x):m\in\M\}$ measures the number of bits
  required to describe $x$ using the best model in the model class.
\end{itemize}
\end{definition}
We can now define which of these representations express the data most efficiently:
\begin{definition}[Candidate set]
The \emph{candidate models} for $x$ from model class $\M$ and threshold function $\tau$ are
\[
  C^\M_\tau(x):=\{m\in\M:L^m(x)\le \tau(x) + c\}
\]
with $c$ a constant independent of $x$ which may be chosen arbitrarily. The default threshold function is $\tau^\circ(x)=L^\M(x)$. We will often omit the subscript if the default threshold function is used. 
\end{definition}

\begin{definition}[Sophistication]
  The model complexity, meaningful information, effective complexity, facticity or \emph{sophistication},  of a string $x$ given a model class $\M$ and threshold function $\tau$ is
  \[
  \s_\tau^{\M}(x):=\min\{L(m):m\in C^\M_\tau(x)\}\p
  \]\label{definition:sophistication}
\end{definition}

Finally, we must formalize what we mean when we say that we desire sophistication to be invariant. For the Kolmogorov complexity, we know that a change in acceptable numbering causes no more than a constant change. It may be that a less stringent invariance (for instance a $\log$ term) can still yield a worthwhile definition of sophistication. To allow such distinctions, we generalize the notion of invariance:

\begin{definition}
Let $\gamma:\N\to \N$ be some nondecreasing function. Two functions $f$ and $g$ are $\gamma$-equivalent if $|f(x) - g(x)| \in O(\gamma(\max(f(x), g(x)))$.
\end{definition}

We say that a formalization of sophistication $s_\phi$ built on acceptable numbering $\phi$ is $\gamma$-invariant to the choice of numbering is $s_\phi$ and $s_\psi$ are $\gamma$-equal for any two acceptable numberings $\phi$ and $\psi$. Thus, by this definition, Kolmogorov complexity is $1$-invariant.

\section{Inefficient indices}

We start from a simple intuition: the definition of Kolmogorov complexity already uses a description of the data in two parts; a computable function and and input. Can we assume that when such a description is as short as the Kolmogorov complexity, it automatically creates a good separation into the structural part of the data and the noise? As we will show, the representation of the structural part plays a fundamental part in answering this question. If it is in any way inefficient, the resulting sophistication function does not behave as expected.

The issues can be very subtle and require a careful attention that is not always afforded in the literature. We will first illustrate what happens if any acceptable numbering is allowed:

@@

We now define two new acceptable numberings $\num{\psi}$ and $\num{\xi}$:
\[
\psi_j(x) = 
\begin{cases}
	\phi_i(x) \;\text{if}\; j = s_{2i}\\
	\infty\;\text{otherwise}
\end{cases}
\]
and 
\[
\xi_j(x) = 
\begin{cases}
	\phi_i(x) \;\text{if}\; j = s_{2i+1} \\
	\infty \;\text{otherwise}\p
\end{cases}
\]
Let $\s_\psi$ and $\s_\xi$ be the sophistication functions built on these enumerations. Choose any $x$ and assume w.l.o.g. that $\s_\psi(x) < \s_\xi(x)$. By construction of the numberings $\psi$ and $\xi$, we have $2\;\s_\psi(x)  \leq \s_\xi(x)$, which gives us:
\begin{align*}
	\s_\xi(x) - \s_\psi(x) &\geq \frac{1}{2}\, \s_\xi(x)  \in \Omega(\s_\xi(x))\p 
\end{align*}
Since, by assumption, $\s_\xi$ can be arbitrarily large and $\s_\xi(x) - \s_\psi(x) \in O(\gamma(\s_\xi(x))) \subset o(\s_\xi(x))$, we have a contradiction.
\end{proof}

In this proof, we deliberately used ridiculously inefficient index functions. Obviously, one could simply pick a more efficient index; the point is that many definitions of sophistication do not mention any such restriction. As we shall see, any inefficiency in the index will lead to a lack of invariance in sophistication.

@@

Interestingly, using an inefficient index is no problem in the definition of Kolmogorov complexity, since we can select as a model a universal Turing machine with an efficient index at only a constant penalty (by the Invariance Theorem \cite{li1993introduction}). But this robustness property does \emph{not} carry over to definitions of sophistication. If we are interested in separating the model and its input, we must represent the model efficiently.

Another perspective on this problem is that the length of a model in an ordinary description does not represent its complexity, much like the length of a compressible string does not reflect its information content. In order to obtain a more robust measure of model complexity, we describe a natural extension of Kolmogorov complexity to functions, and show that this retains invariance.

\begin{definition}
  The complexity of a partial recursive function is defined by
  \begin{align*}
    C_\phi(f) &:= \min\{C(i):\phi_i=f\}\\
    K_\phi(f) &:= \min\{K(i):\phi_i=f\}.
  \end{align*} \label{definition:model-complexity}
\end{definition}
These definitions also appear in \cite{grunwald2004shannon} and \cite{vitanyi2004meaningful}. We will show that these values are invariant up to a constant (ie. complexity is an objective property of a function):
\begin{lemma}[Invariance]
Let $\phi_1, \phi_2, \ldots$ and $\psi_1, \psi_2,\ldots$ be two acceptable numberings. There exists a constant $c$ such that $\left| K^\phi(f) - K^\psi(f)\right | \leq c$ for all $f$. \label{lemma:invariance}
\end{lemma}
\begin{proof}
Let $g(i)$ be the function which which translates from $\psi$ to $\phi$.
\begin{align*}
K^\phi(f) &= \min\left\{ K^\phi(i) : \phi_i= f\right\} \\
&\geq \min\left\{ K^\psi(i) : \phi_i= f\right\} - c\\
&= \min\left\{ K^\psi(i) : \psi_{g(i)}= f\right\} - c\\
&\geq \min\left\{ K^\psi(i) : \psi_i= f\right\} - c\\
&= K^\psi(f).
\end{align*}
We can reverse $\phi$ and $\psi$ to achieve the same inequality the other way around, completing the proof.
\end{proof}
This lemma shows that the complexity of a function does not depend on the choice of
enumeration by more than a constant term.

>>>>

\newcommand{\M}{\mathscr M}
\newcommand{\C}{\mathscr C}
\newcommand{\T}{\mathscr T}
\newcommand{\F}{\mathscr F}
\renewcommand{\P}{\mathscr P}
\newcommand{\K}{\mathscr K}
\newcommand{\X}{\mathscr X}
\newcommand{\B}{\mathbb B}
\newcommand{\D}{\Delta}
\newcommand{\N}{\mathbb N}
\newcommand{\tn}[1]{\textnormal{#1}}
\newcommand{\pair}[1]{\left\langle{#1}\right\rangle}
\newcommand{\concat}{\oplus}
\newcommand{\symb}[1]{\texttt{#1}}
\newcommand{\br}[1]{\overline{#1}}
\newcommand{\s}{S}
\newcommand{\num}[1]{#1_1, #1_2, \ldots}

@@

\begin{abstract}
Kolmogorov complexity is a sound measure of the amount of information in data, but it does not distinguish structural information from random noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to capture this separation. Since then many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. The multitude of treatments indicates that the idea is considered worth pursuing by multiple distinguished scholars, but at the same time it shows that the issue has never been satisfactorily resolved. We describe three fundamental problems with all existing proposals, showing each of them to be \emph{unsound}. Based on these results we we put forward the view that the problem is fundamental: while many nontrivial and useful structural properties of binary strings can be detected without a priori assumptions, it may be impossible to objectively quantify the sophistication.
\end{abstract}

@@

Kolmogorov complexity breaks away from the tradition in statistics and information theory by allowing the analysis of \emph{individual} binary sequences rather than studying large samples from probability distributions in an assumed class. This is important because for many binary strings, such as a binary encoding of a novel, or a holiday snapshot, or the fossil record, it is impossible to determine what distribution generates such objects. Kolmogorov complexity is often seen as the first step to a statistics without assumptions.

There is, however, a nagging discrepancy between Kolmogorov complexity and what most people would call ``complex''. For example, a sequence of a million coin flips is exceedingly unlikely to have a Kolmogorov complexity of much less than one million bits, even though there is nothing complex about flipping a lot of coins---it's just hard work. For this reason, many scholars have defined alternative complexity measures in the spirit of Kolmogorov complexity, aimed at quantifying not \emph{all} information in a binary string, but only the \emph{meaningful} information in the data. This tradition started with a lecture by Kolmogorov in Talinn in 1973, but many alternative definitions of meaningful information, or \emph{sophistication},\footnotemark were proposed, see the related work section below. In this paper, we investigate three problems with these definitions, and show that they are unsatisfactory. We believe the problem is fundamental, and provide several intuitive arguments why these attempts to define sophistication must fail.

\footnotetext{Most proposals coin their own name along with a specific treatment. For the sake of clarity, we have chosen to use \emph{sophistication} as an umbrella term to refer to all ideas that fit this mold.}

The Kolmogorov complexity of a binary string $x$ is, roughly, the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings. We can see it as \emph{a property of the data}.

@@

\begin{enumerate}
\item Since it measures information, the sophistication should count the number of bits required for an effective description of the structural properties of a binary string.
\item An analogue of the Invariance Theorem should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language, even for inputs of very high Kolmogorov complexity. Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $K(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}

For most proposed definitions of sophistication, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only extremely deep strings (i.e., strings that require an enormous amount of processing to construct) yield a large sophistication, while intuitively it seems there must be objects that contain a lot of structure all of which can be described without much calculation. Moreover in our discussion section we provide arguments that suggest that it should be possible to generalise our negative results to show that Vit\'anyi's sophistication is not invariant.

A valid definition of sophistication must contend with the following issues. First, while all definitions are based on a two-part encoding of the data sequence, some researchers use Kolmogorov complexity as a \emph{building block} of their proposed code, while others look for the two parts of the code \emph{inside} the Kolmogorov complexity: when the definition of Kolmogorov complexity is opened up, it turns out to consist of two parts itself. In Section~\ref{section:indices}, we show how this second approach sets us up to suffer from to the so-called \emph{nickname problem}: the sophistication becomes highly dependent on the chosen programming language. This immediately breaks invariance, unless additional requirements are made on the language. We also show that there do exist languages (efficient enumerations of the partial recursive functions) that avoid this issue.

The second issue is that of \emph{overfitting}, which we consider in Section~\ref{section:overfitting}. Overfitting is a common problem in the statistical literature, that refers to the tendency of some model selection procedures to select a very complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. For example, in a polynomial regression setting, a higher order polynomial can always provide better fit to a data set. Therefore, even if the data are in fact noisy samples from a parabola, naive model selection methods may select a much higher order polynomial that provides a worse description of the trend in the data, encoding all noise in the model parameters. The same can occur for the sophistication. We somehow need to make sure that the two-part representation that determines the sophistication does not store noise in the ``structure'' part of the code. In statistics, this is typically addressed by penalising complex models. However, in the sophistication setting, such penalties tend to break the delicate balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting (Section~\ref{section:underfitting}) is the opposite problem, where the selected model is simple, but fails to represent part of the structure in the data. It is an especially serious problem for sophistication because the models for used in sophistication are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such interpreters, or \emph{universal models}, can be described in a constant amount of bits, and can generate the data with from a program the length of the Kolmogorov complexity, essentially encoding all information as noise. If complex models are penalized to avoid overfitting, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions (only programs that halt for all inputs) rather than all partial recursive functions (all programs). While this patches the problem of universal models to an extent, it is questionable whether it adequately solves the problem of underfitting in general.

Finally, in the discussion in Section~\ref{section:conclusion} we argue that while two-part coding can yield useful insights into the structure of the data and allows us to identify some models as unlikely to capture all structure in the data, it is probably not possible to uniquely separate structure from noise and identify a \emph{single} model as ``best'': in general many models, of vastly different complexities, may be reasonable models of the data generating mechanism. Rather than doggedly trying to ``fix'' this property of algorithmic statistics, we propose embracing the idea that sometimes the data allows for multiple equivalent interpretations, and that there is no such thing as sophistication.

\subsection{Related work}
The idea of separating structure from noise originated with Kolmogorov's lecture in Tallinn, Estonia in 1973 introducing the \emph{structure function} \cite{cover1985kolmogorov}. In this theory, the two-part description of the data consists of a program that generates a finite set, and an index of the original object in the set. The length of the program then measures the structure in the data. This approach was discussed more extensively, and generalised, by Vereshchagin, G\'acs and Vit\'anyi \cite{vereshchagin2004kolmogorov,gacs2001algorithmic}.

@@

In 1996 Gell-Mann and Lloyd introduced \emph{effective complexity} \cite{gellmann1996information}. Working from the perspective of statistical mechanics, they arrived (largely independently, it seems) at a notion that fits the mold of sophistication.

Antunes et al. \cite{antunes2009sophistication} note that the constant cutoff used by Koppel to separate candidate two-part descriptions from those deemed to be too inefficient, is a source of instability, a result in the spirit of this paper. They propose a different definition called \emph{coarse sophistication}. As we shall see, this version behaves very differently from regular sophistication, and is closer to computational depth. Vereshchagin \cite{vereshchagin2013algorithmic} proposes an alternative solution to the instability issue.

Adriaans' \emph{Facticity}, introduced in 2012\cite{adriaans2012facticity}, relaxes the model class to include all partial recursive functions. Adriaans also identifies the nickname problem discussed in Section~\ref{section:indices}.

@@

@@

\begin{definition}[Numberings]
A numbering is an enumeration of of the partial computable functions, denoted as $\phi_1, \phi_2, \ldots$. We fix one canonical, which we know to be an effective description, that is for any $i \in \N, y\in \B$, we can effectively compute $\phi_i(y)$, where the computation diverges if $\phi_i(y) = \infty$. We call a numbering $\num{\psi}$ \emph{acceptable} if there exist total, computable functions $a, b: \N \to \N$ such that for all $i$, $\phi_i = \psi_{b(i)}$ and  $\psi_i = \phi_{b(i)}$. We use the greek letters without a subscript (eg. $\phi$ and $\psi$) to refer to the numbering as a whole.
\end{definition}

@@

\begin{definition}[Complexity]\belowdisplayskip=-12pt
Let $\phi_1,\phi_2,\ldots$ be the canonical acceptable enumeration of $\C$. Let $\Pi$ be a computable subset of the natural numbers such that  $\{\phi_i:i\in\Pi\}$ corresponds to the set of all computable prefix functions. We define the plain and prefix Kolmogorov complexity as 
\begin{align*}
C(x)&=\min\{|\bar\imath y|:\phi_i(y)=x\},\\
K(x)&=\min\{|\bar\imath y|:\phi_i(y)=x, i\in\Pi\}.\\
\end{align*}\label{definition:complexity}\vspace{-18pt}
\end{definition}
This definition of $K$ is uncommon. It is equivalent to the standard definition as used in \cite{li1993introduction}, but this form is more convenient for defining the $K$-complexity of non-prefix functions (see Definition~\ref{definition:model-complexity} below).

We will generalize the main proposals for sophistication into a single notation. We start with a class of models, we consider all two-part descriptions of the data using these models, we select the set of representations with the shortest code-lengths and we select the smallest model from this set. The size of the smallest model represents the sophistication of the data.

\begin{definition}[Models and model classes] A \emph{model class} is a set of models. We distinguish four important model classes:
  \begin{itemize}
  \item $\C$ is the set of all partial computable functions,
  \item $\K\subset\C$ is the set of all prefix-free
    functions,
  \item $\T\subset\C$ is the set of all total functions,
  \item $\P\subset\K$ is the set of all complete, prefix-free functions,
  \item $\F\subset\P$ contains, for every finite set $S$, a computable
    surjective function mapping the binary sequences of length
    $\lceil\log S\rceil$ onto $S$. In other words, this is the model class of finite sets.
  \end{itemize}
\end{definition}
We note that classes $\T$ and $\P$ are unusual in that they are not enumerable.
\begin{definition}[Representations and code-lengths]
A pair $(m,y)$ is a \emph{representation} for $x$ if $m$ is a model and $m(y)=x$.
Define the following code-length functions for the representations:
\begin{itemize}
\item $L(m)$ is the length of a self-delimiting description of
  the model $m$, i.e. it is the length function of a prefix code for models. 
\item $L^m(x)=L(m)+\min\{|y|:m(y)=x\}$ measures the number of bits required to describe both the model $m$ and the data $x$ using the model.
\item $L^\M(x)=\min\{L^m(x):m\in\M\}$ measures the number of bits
  required to describe $x$ using the best model in the model class.
\end{itemize}
\end{definition}
We can now define which of these representations express the data most efficiently:
\begin{definition}[Candidate set]
The \emph{candidate models} for $x$ from model class $\M$ and threshold function $\tau$ are
\[
  C^\M_\tau(x):=\{m\in\M:L^m(x)\le \tau(x) + c\}
\]
with $c$ a constant independent of $x$ which may be chosen arbitrarily. The default threshold function is $\tau^\circ(x)=L^\M(x)$. We will often omit the subscript if the default threshold function is used. 
\end{definition}

\begin{definition}[Sophistication]\belowdisplayskip=-12pt
  The model complexity, meaningful information, effective complexity, facticity or \emph{sophistication},  of a string $x$ given a model class $\M$ and threshold function $\tau$ is
  \[
  \s_\tau^{\M}(x):=\min\{L(m):m\in C^\M_\tau(x)\}\p
  \]\label{definition:sophistication}
\end{definition}
Finally, we must formalize what we mean when we say that we expect sophistication to be invariant. It may be that a less strict invariance than we have for the Kolmogorov complexity, still yields a useful sophistication.
\begin{definition}
Let $\gamma:\N\to \N$ be some nondecreasing function. Two functions $f$ and $g$ are $\gamma$-equivalent if $|f(x) - g(x)| \in O(\gamma(\max(f(x), g(x)))$.
\end{definition}
We say that a formalization of sophistication $\s_\phi$ built on acceptable numbering $\phi$ is $\gamma$-invariant to the choice of numbering is $s_\phi$ and $s_\psi$ are $\gamma$-equal for any two acceptable numberings $\phi$ and $\psi$. 
\begin{lemma}
If a function $\S_\phi(x)$ is bounded for a given numbering, it is either not $\gamma$-invariant or it is  $\gamma$-equal to a bounded function.
\end{lemma}
The proof follows directly from the definitions. Most lemmas in the rest of the paper will simply prove that a given variant is bounded for a specific numbering, implicitly using this lemma to show that the variant is unsound.

\section{Inefficient indices}
\label{section:indices}
We start from a simple intuition: the definition of Kolmogorov complexity already uses a description of the data in two parts; a computable function and an input. Can we assume that, when such a description is as short as the Kolmogorov complexity, it automatically creates a good separation into the structural part of the data and the noise? As we will show, the representation of function plays a fundamental part in answering this question.

The issues can be very subtle and require a careful attention that is rarely given in the literature. We will first illustrate what happens if any acceptable numbering is allowed:

@@

We now define two new acceptable numberings $\num{\psi}$ and $\num{\xi}$:
\[
\psi_j(x) = \begin{cases}
	\phi_i(x) \;\text{if}\; j = s_{2i}\\
	\infty\;\text{otherwise}
\end{cases}\,\,\,
\xi_j(x) = 
\begin{cases}
	\phi_i(x) \;\text{if}\; j = s_{2i+1} \\
	\infty \;\text{otherwise}\p
\end{cases}
\]
Let $\s_\psi$ and $\s_\xi$ be the sophistication functions built on these enumerations. Choose any $x$ and assume w.l.o.g. that $\s_\psi(x) < \s_\xi(x)$. By construction of the numberings $\psi$ and $\xi$, we have $2\;\s_\psi(x)  \leq \s_\xi(x)$, which gives us:
\begin{align*}
	\s_\xi(x) - \s_\psi(x) &\geq \frac{1}{2}\, \s_\xi(x)  \in \Omega(\s_\xi(x))\p 
\end{align*}
Since, by assumption, $\s_\xi$ can be arbitrarily large and $\s_\xi(x) - \s_\psi(x) \in O(\gamma(\s_\xi(x))) \subset o(\s_\xi(x))$, we have a contradiction.
\end{proof}
In this proof, we deliberately used ridiculously inefficient index functions. Obviously, one could simply pick a more efficient index; the point is that many definitions of sophistication do not mention any such restriction. As we shall see, any inefficiency in the index will lead to a lack of invariance in sophistication.

@@

Interestingly, using an inefficient index is no problem in the definition of Kolmogorov complexity, since we can select as a model a universal Turing machine with an efficient index at only a constant penalty. But this robustness does \emph{not} carry over to definitions of sophistication. If we are interested in separating the model and its input, we must represent the model efficiently.

Simply put, the length of a model in an ordinary description does not represent its complexity, much like the length of a compressible string does not reflect its information content. In order to obtain a more robust measure of model complexity, we describe a natural extension of Kolmogorov complexity to functions, and show that this retains invariance.

\begin{definition}\belowdisplayskip=-12pt
  The complexity of a partial recursive function is defined by
  \begin{align*}
    C_\phi(f) &:= \min\{C(i):\phi_i=f\}\;\tn{,}\\
    K_\phi(f) &:= \min\{K(i):\phi_i=f\}\p
  \end{align*} \label{definition:model-complexity}
\end{definition}
These definitions also appear in \cite{grunwald2004shannon} and \cite{vitanyi2004meaningful}. We will show that these values are invariant up to a constant (ie. complexity is an objective property of a function):
\begin{lemma}[Invariance]
Let $\phi_1, \phi_2, \ldots$ and $\psi_1, \psi_2,\ldots$ be any two acceptable numberings. There exists a constant $c$ such that $\left| K_\phi(f) - K_\psi(f)\right | \leq c$ for all $f$. \label{lemma:invariance}
\end{lemma}
\begin{proof}
Let $g(i)$ be the function such that $\psi_i$ to $\phi_{g(i)}$.
\begin{align*}
K_\phi(f) &= \min\left\{ K_\phi(i) : \phi_i= f\right\} 
\geq \min\left\{ K_\psi(i) : \phi_i= f\right\} - c\\
&= \min\left\{ K_\psi(i) : \psi_{g(i)}= f\right\} - c
\geq \min\left\{ K_\psi(i) : \psi_i= f\right\} - c = K_\psi(f).
\end{align*}
The opposite inequality can be achieved by reversing $\phi$ and $\psi$. 
\end{proof}
This lemma shows that the complexity of a function does not depend on the choice of
enumeration by more than a constant term.
